---
title: "practical_exercise_5, Methods 3, 2021, autumn semester"
author: '[FILL IN YOUR NAME]'
date: "[FILL IN THE DATE]"
output: pdf_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, lmerTest, lme4, gridExtra, dfoptim, readbulk)
```

# Exercises and objectives
The objectives of the exercises of this assignment are based on: https://doi.org/10.1016/j.concog.2019.03.007  
  
4) Download and organise the data from experiment 1  
5) Use log-likelihood ratio tests to evaluate logistic regression models  
6) Test linear hypotheses  
7) Estimate psychometric functions for the Perceptual Awareness Scale and evaluate them  

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below (__MAKE A KNITTED VERSION__)  
REMEMBER: This is part 2 of Assignment 2 and will be part of your final portfolio


# EXERCISE 4 - Download and organise the data from experiment 1

Go to https://osf.io/ecxsj/files/ and download the files associated with Experiment 1 (there should be 29).  
The data is associated with Experiment 1 of the article at the following DOI https://doi.org/10.1016/j.concog.2019.03.007  
  
1) Put the data from all subjects into a single data frame - note that some of the subjects do not have the _seed_ variable. For these subjects, add this variable and make in _NA_ for all observations. (The _seed_ variable will not be part of the analysis and is not an experimental variable)  
    i. Factorise the variables that need factorising  
    ii. Remove the practice trials from the dataset (see the _trial.type_ variable)  
    iii. Create a _correct_ variable  
    iv. Describe how the _target.contrast_ and _target.frames_ variables differ compared to the data from part 1 of this assignment
    
```{r}

df = read_bulk("experiment_1")
#read_bulk() automatically fills empty rows with NA, so no need to do anything extra

df = df %>% 
  filter(trial.type == "experiment")
df = df %>%
  mutate(correct = ifelse(target.type == "even" & obj.resp == "e" |
                          target.type == "odd" & obj.resp == "o", 1, 0))

# Assigning variables to proper class
df$pas <- as.factor(df$pas)
df$trial <- as.character(df$trial)
df$target.contrast <- as.numeric(df$target.contrast)
df$cue <- as.character(df$cue)
df$rt.subj <- as.numeric(df$rt.subj)
df$rt.obj <- as.numeric(df$rt.obj)
df$target.frames <- as.integer(df$target.frames)
df$correct <- as.numeric(df$correct)
df$subject <- as.factor(df$subject)

```

**ALEKS**: *Target.contrast* is not manipulated in this experiment and is set at 0.1. *target.frames* is now manipulated and ranges from 1-6 frames, with one frame being 11.8ms.
 
# EXERCISE 5 - Use log-likelihood ratio tests to evaluate logistic regression models

1) Do logistic regression - _correct_ as the dependent variable and _target.frames_ as the independent variable. (Make sure that you understand what _target.frames_ encode). Create two models - a pooled model and a partial-pooling model. The partial-pooling model should include a subject-specific intercept. 

```{r}

#Pooled model
m1 = glm(correct ~ target.frames, df, family = "binomial")

#Partial pooled model
m2 = glmer(correct ~ target.frames + (1|subject), df, family = "binomial")
summary(m2)

```


    i. the likelihood-function for logistic regression is: $L(p)={\displaystyle\prod_{i=1}^Np^{y_i}(1-p)^{(1-y_i)}}$ (Remember the probability mass function for the Bernoulli Distribution). Create a function that calculates the likelihood.  
    
```{r}

likelihood_fun = function(i) {
  p = fitted(i) # Vector of fitted values
  y = as.vector(model.response(model.frame(i), type = "numeric")) # Observed y-values
  likelihood = prod(p^y*(1-p)^(1-y)) # The likelihood function for logistic regression
  return(likelihood)
}

likelihood_fun(m2)

```

**ALEKS**: The likelihood function outputs 0 because it's so close to zero R treats it as such.

    
    ii. the log-likelihood-function for logistic regression is: $l(p) = {\displaystyle\sum_{i=1}^N}[y_i\ln{p}+(1-y_i)\ln{(1-p)}$. Create a function that calculates the log-likelihood  
    

```{r}

log_likelihood_fun = function(i) {
  p = fitted(i) # Vector of fitted values
  y = as.vector(model.response(model.frame(i), type = "numeric")) # Observed y-values
  log_likelihood = sum(y*log(p)+(1-y)*log(1-p)) # The log-likelihood function for logistic regression
  return(log_likelihood)
}

log_likelihood_fun(m2)

```

    
    iii. apply both functions to the pooling model you just created. Make sure that the log-likelihood matches what is returned from the _logLik_ function for the pooled model. Does the likelihood-function return a value that is surprising? Why is the log-likelihood preferable when working with computers with limited precision?  
    
```{r}

likelihood_fun(m1)
log_likelihood_fun(m1)
logLik(m1)

```

**ALEKS**: Output is the same for both functions. The log-likelihood is better, because the likelihood function has to calculate more decimals (or the decimals are more meaningful, because the likelihood will often approximate 0).
    
    iv. now show that the log-likelihood is a little off when applied to the partial pooling model - (the likelihood function is different for the multilevel function - see section 2.1 of https://www.researchgate.net/profile/Douglas-Bates/publication/2753537_Computational_Methods_for_Multilevel_Modelling/links/00b4953b4108d73427000000/Computational-Methods-for-Multilevel-Modelling.pdf if you are interested)  
    
```{r}

log_likelihood_fun(m2)
logLik(m2)

```

    
2) Use log-likelihood ratio tests to argue for the addition of predictor variables, start from the null model, `glm(correct ~ 1, 'binomial', data)`, then add subject-level intercepts, then add a group-level effect of _target.frames_ and finally add subject-level slopes for _target.frames_. Also assess whether or not a correlation between the subject-level slopes and the subject-level intercepts should be included.

```{r}

m0 = glm(correct ~ 1, data = df, family = 'binomial') # Null-model
m01 = glmer(correct ~ 1 + (1|subject), data = df, family = 'binomial') # Null-model with subject intercepts
#m2 = Model from before, predicted by target.frames and subject intercepts
m3 = glmer(correct ~ target.frames + (target.frames|subject), data = df, family = "binomial")
anova(m01, m0, m2, m3) # This function also give logLik values
anova(m2, m3)

```


    i. write a short methods section and a results section where you indicate which model you chose and the statistics relevant for that choice. Include a plot of the estimated group-level function with `xlim=c(0, 8)` that includes the estimated subject-specific functions.
    
**ALEKS**: I use the anova function to display the logLik values of several models with varying complexity. Since m3 and m2 have the highest logLik values (most likely to be true), I compare these to see if the difference between them is significant. The difference between the two models is significant (p < 0.05), which is why i would choose the more complex model with target.frames as a fixed effect and with intercepts per subject and slopes for target.frames. **a**
    
    ii. also include in the results section whether the fit didn't look good for any of the subjects. If so, identify those subjects in the report, and judge (no statistical test) whether their performance (accuracy) differed from that of the other subjects. Was their performance better than chance? (Use a statistical test this time) (50 %)  
    
```{r}

df_24 = df %>%
  filter(subject == "24")

t.test(df_24$correct, mu = 0.5)

```
**ALEKS**: When running a one sample t-test which the null hypothesis of "true mean is equal to 0.5", we obtain a p-value of less then 0.05, meaning that we reject the null hypothesis. According to this, subject 24 performs better than chance. **a**
    
3) Now add _pas_ to the group-level effects - if a log-likelihood ratio test justifies this, also add the interaction between _pas_ and _target.frames_ and check whether a log-likelihood ratio test justifies this  

```{r}

m4 = glmer(correct ~ target.frames + pas + (target.frames|subject), data = df, family = "binomial")
m5 = glmer(correct ~ target.frames * pas + (target.frames|subject), data = df, family = "binomial")

anova(m3, m4, m5)
anova(m4, m5)

```


    i. if your model doesn't converge, try a different optimizer  
    
**ALEKS**: All good it does. My M1 (not to be confused with m1 model) Mac is so fast.

    ii. plot the estimated group-level functions over `xlim=c(0, 8)` for each of the four PAS-ratings - add this plot to your report (see: 5.2.i) and add a description of your chosen model. Describe how _pas_ affects accuracy together with target duration if at all. Also comment on the estimated functions' behaviour at target.frame=0 - is that behaviour reasonable?  
    
```{r}

df %>% # Plot of group-level function per subject
  ggplot() +
   geom_point(aes(x = target.frames, y = fitted(m5), color = pas)) + 
   facet_wrap( ~ pas) + 
   labs(title = "Estimated group-level function pr. subject") +
   theme_bw()

```
**ALEKS**: The plot shows that for the lowest PAS score of 1, increasing target frames both increases accuracy for some subjects and reduces it for others. When PAS score increases, we can also see that the distribution of fitted values is reduced, and at PAS rating 4, large value of target.frames approximates almost 100% probability of correctness. *a*

# EXERCISE 6 - Test linear hypotheses

In this section we are going to test different hypotheses. We assume that we have already proved that more objective evidence (longer duration of stimuli) is sufficient to increase accuracy in and of itself and that more subjective evidence (higher PAS ratings) is also sufficient to increase accuracy in and of itself. 

We want to test a hypothesis for each of the three neighbouring differences in PAS, i.e. the difference between 2 and 1, the difference between 3 and 2 and the difference between 4 and 3. More specifically, we want to test the hypothesis that accuracy increases *faster* with objective evidence if subjective evidence is *higher at the same time*, i.e. we want to test for an interaction.  

1) Fit a model based on the following formula: `correct ~ pas * target.frames + (target.frames | subject))`
    i. First, use `summary` (yes, you are allowed to!) to argue that accuracy increases faster with objective evidence for PAS 2 than for PAS 1. 
    
```{r}

summary(m5)

```


2) `summary` won't allow you to test whether accuracy increases faster with objective evidence for PAS 3 than for PAS 2 (unless you use `relevel`, which you are not allowed to in this exercise). Instead, we'll be using the function `glht` from the `multcomp` package
    i. To redo the test in 6.1.i, you can create a _contrast_ vector. This vector will have the length of the number of estimated group-level effects and any specific contrast you can think of can be specified using this. For redoing the test from 6.1.i, the code snippet below will do
    ii. Now test the hypothesis that accuracy increases faster with objective evidence for PAS 3 than for PAS 2.
    iii. Also test the hypothesis that accuracy increases faster with objective evidence for PAS 4 than for PAS 3
3) Finally, test that whether the difference between PAS 2 and 1 (tested in 6.1.i) is greater than the difference between PAS 4 and 3 (tested in 6.2.iii)

### Snippet for 6.2.i
```{r, eval=FALSE}
## testing whether PAS 2 is different from PAS 1
contrast.vector <- matrix(c(0, 0, 0, 0, 0, 1, 0, 0), nrow=1)
gh <- glht(pas.intact.tf.ranslopeint.with.corr, contrast.vector)
print(summary(gh))
## as another example, we could also test whether there is a difference in
## intercepts between PAS 2 and PAS 3
contrast.vector <- matrix(c(0, -1, 1, 0, 0, 0, 0, 0), nrow=1)
gh <- glht(pas.intact.tf.ranslopeint.with.corr, contrast.vector)
print(summary(gh))
```

# EXERCISE 7 - Estimate psychometric functions for the Perceptual Awareness Scale and evaluate them  

We saw in 5.3 that the estimated functions went below chance at a target duration of 0 frames (0 ms). This does not seem reasonable, so we will be trying a different approach for fitting here.  
We will fit the following function that results in a sigmoid, $f(x) = a + \frac {b - a} {1 + e^{\frac {c-x} {d}}}$  
It has four parameters: _a_, which can be interpreted as the minimum accuracy level, _b_, which can be interpreted as the maximum accuracy level, _c_, which can be interpreted as the so-called inflexion point, i.e. where the derivative of the sigmoid reaches its maximum and _d_, which can be interpreted as the steepness at the inflexion point. (When _d_ goes towards infinity, the slope goes towards a straight line, and when it goes towards 0, the slope goes towards a step function).  
  
We can define a function of a residual sum of squares as below

```{r, eval=FALSE}
RSS <- function(dataset, par)
{
    ## "dataset" should be a data.frame containing the variables x (target.frames)
    ## and y (correct)
    
    ## "par" are our four parameters (a numeric vector) 
    a = par[1] 
    b = par[2]
    c = par[3] 
    d = par[4]
    
    x <- dataset$x
    y <- dataset$y
    
    y.hat <- a + ((b-a)/(1+exp(1)^((c-x)/d))) 
    
    RSS <- sum((y - y.hat)^2)
    return(RSS)
}
```

1) Now, we will fit the sigmoid for the four PAS ratings for Subject 7
    i. use the function `optim`. It returns a list that among other things contains the four estimated parameters. You should set the following arguments:  
    `par`: you can set _c_ and _d_ as 1. Find good choices for _a_ and _b_ yourself (and argue why they are appropriate)  
    `fn`: which function to minimise?  
    `data`: the data frame with _x_, _target.frames_, and _y_, _correct_ in it  
    `method`: 'L-BFGS-B'  
    `lower`: lower bounds for the four parameters, (the lowest value they can take), you can set _c_ and _d_ as `-Inf`. Find good choices for _a_ and _b_ yourself (and argue why they are appropriate)  
    `upper`: upper bounds for the four parameters, (the highest value they can take) can set _c_ and _d_ as `Inf`. Find good choices for _a_ and _b_ yourself (and argue why they are appropriate)    
    ii. Plot the fits for the PAS ratings on a single plot (for subject 7) `xlim=c(0, 8)`
    iii. Create a similar plot for the PAS ratings on a single plot (for subject 7), but this time based on the model from 6.1 `xlim=c(0, 8)`   
    iv. Comment on the differences between the fits - mention some advantages and disadvantages of each way  
2) Finally, estimate the parameters for all subjects and each of their four PAS ratings. Then plot the estimated function at the group-level by taking the mean for each of the four parameters, _a_, _b_, _c_ and _d_ across subjects. A function should be estimated for each PAS-rating (it should look somewhat similar to Fig. 3 from the article:  https://doi.org/10.1016/j.concog.2019.03.007)
    i. compare with the figure you made in 5.3.ii and comment on the differences between the fits - mention some advantages and disadvantages of both.
    
```{r}
# Subsetting df
df_7 <- df %>% 
  filter(subject == "7") %>% 
  dplyr::select(target.frames, correct, pas) %>% 
  rename(x = target.frames, y = correct)
# Specifying par-values
par <- c(0.5, 1, 1, 1)
# Running the optim function for each pas score
optim_7_pas1 <- optim(data = filter(df_7, pas == "1"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
optim_7_pas2 <- optim(data = filter(df_7, pas == "2"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
optim_7_pas3 <- optim(data = filter(df_7, pas == "3"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
optim_7_pas4 <- optim(data = filter(df_7, pas == "4"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
# Showing parameter estimates pr PAS rating
print(optim_7_pas1)
print(optim_7_pas2)
print(optim_7_pas3)
print(optim_7_pas4)
```

Argument for a and b values: The lowest accuracy we would expect is 50%, as this is the accuracy we would acquire from randomly drawing. The maximum accuracy would be 100%, if correctness was perfectly predicted from every input value.

### 7.1.ii. Plot the fits for the PAS ratings on a single plot (for subject 7) xlim=c(0, 8)

```{r}

# Defining a sigmoid-function that takes parameters from optim-function and x-values from a df.
sigmoid_function <- function(optimfun, x) {
    optimfun$par[1] + ((optimfun$par[2]-optimfun$par[1])/(1+exp(1)^((optimfun$par[3]-x)/optimfun$par[4])))}
# Adding y_hats to dataframe
df_7$y_hat_pas1 <- sigmoid_function(optim_7_pas1,df_7$x)
df_7$y_hat_pas2 <- sigmoid_function(optim_7_pas2,df_7$x)
df_7$y_hat_pas3 <- sigmoid_function(optim_7_pas3,df_7$x)
df_7$y_hat_pas4 <- sigmoid_function(optim_7_pas4,df_7$x)
# plotting 
df_7 %>% 
  ggplot() + 
  geom_smooth(aes(x = x, y = y_hat_pas1, color = "pas1")) + 
  geom_smooth(aes(x = x, y = y_hat_pas2, color = "pas2")) + 
  geom_smooth(aes(x = x, y = y_hat_pas3, color = "pas3")) + 
  geom_smooth(aes(x = x, y = y_hat_pas4, color = "pas4")) +
  labs(title = "Estimated fits for accuracy ratings pr. PAS for subject 7",
       x = "Target.Frames",
       y = "Estimated accuracy ratings using sigmoid-function") +
  theme_bw() 

```

### 7.1.iii. Create a similar plot for the PAS ratings on a single plot (for subject 7), but this time based on the model from 6.1 xlim=c(0, 8)

```{r}

# Subsetting to only use values from subject 7
df_7_estimates <- df
df_7_estimates$fitted <- fitted(m5)
df_7_estimates <- df_7_estimates %>% 
  filter(subject == "7")
# Plotting (NO OBSERVATIONS OF TARGET.FRAMES = 0 AND PAS = 4)
df_7_estimates %>% 
  ggplot() + 
  geom_line(aes(x = target.frames, y = fitted, color = pas))+
  labs(title = "Estimated accurary from interaction-model (for subject 7)") +
  labs(x = "Target duration (target.frames)", y = "Estimated Accuracy") + 
  theme_bw()

```

### 7.1.iv. Comment on the differences between the fits - mention some advantages and disadvantages of each way

The model created with the glm-function uses the data from all subjects to create a more appropriate model, whereas simply fitting the "best" sigmoid using the optim function for a single subject's data will result in overfitting. The optim-function has the advantage of providing the actual parameters for creating the sigmoid function.

### 7.2. Finally, estimate the parameters for all subjects and each of their four PAS ratings. Then plot the estimated function at the group-level by taking the mean for each of the four parameters, *a*, *b*, *c* and *d* across subjects. A function should be estimated for each PAS-rating (it should look somewhat similar to Fig. 3 from the article: <https://doi.org/10.1016/j.concog.2019.03.007>)

```{r}

pas_rating_function <- function(dataframe, participant){
  
  # Subsetting the df
  dataframe <- dataframe %>% 
    dplyr::filter(subject == participant) %>% 
    dplyr::select(subject, target.frames, correct, pas) %>% 
    dplyr::rename(x = target.frames, y = correct)
  
  # Specifying par
  par <- c(0.5, 1, 1, 1)
  
  optim_pas1 <- optim(data = filter(dataframe, pas == "1"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
  optim_pas2 <- optim(data = filter(dataframe, pas == "2"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
  optim_pas3 <- optim(data = filter(dataframe, pas == "3"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
  optim_pas4 <- optim(data = filter(dataframe, pas == "4"), fn = RSS, par = par, method = 'L-BFGS-B', lower = c(0, 0, -Inf, -Inf), upper = c(1, 1, Inf, Inf))
  
  # Now i have 4 variables for each pas score based on the dataframe and participant
  dataframe$a_value_pas_1 <- optim_pas1$par[1]
  dataframe$b_value_pas_1 <- optim_pas1$par[2]
  dataframe$c_value_pas_1 <- optim_pas1$par[3]
  dataframe$d_value_pas_1 <- optim_pas1$par[4]
  
  # Running the sigmoid-function to get parameter estimates
  dataframe$y_hat_1 <- sigmoid_function(optim_pas1, dataframe$x)
  dataframe$y_hat_2 <- sigmoid_function(optim_pas2, dataframe$x)
  dataframe$y_hat_3 <- sigmoid_function(optim_pas3, dataframe$x)
  dataframe$y_hat_4 <- sigmoid_function(optim_pas4, dataframe$x)
  
  # Getting mean values per x (target.frames)
  dataframe <- dataframe %>% 
    group_by(x) %>% 
    mutate(y_hat_1_mean = mean(y_hat_1),
         y_hat_2_mean = mean(y_hat_2),
         y_hat_3_mean = mean(y_hat_3),
         y_hat_4_mean = mean(y_hat_4)) %>% 
    ungroup()
  
  return(dataframe)
}
# Estimated values loop
new_df <- data.frame()
for (i in 1:29){
  newer_df <- pas_rating_function(df, i)
  new_df <- rbind(new_df, newer_df)
}
# Extracting mean parameters from parameters df (very clunky way to do it)
a_mean_pas_1 <- mean(new_df$a_value_pas_1)
b_mean_pas_1 <- mean(new_df$b_value_pas_1)
c_mean_pas_1 <- mean(new_df$c_value_pas_1)
d_mean_pas_1 <- mean(new_df$d_value_pas_1)
a_mean_pas_2 <- mean(new_df$a_value_pas_2)
b_mean_pas_2 <- mean(new_df$b_value_pas_2)
c_mean_pas_2 <- mean(new_df$c_value_pas_2)
d_mean_pas_2 <- mean(new_df$d_value_pas_2)
a_mean_pas_3 <- mean(new_df$a_value_pas_3)
b_mean_pas_3 <- mean(new_df$b_value_pas_3)
c_mean_pas_3 <- mean(new_df$c_value_pas_3)
d_mean_pas_3 <- mean(new_df$d_value_pas_3)
a_mean_pas_4 <- mean(new_df$a_value_pas_4)
b_mean_pas_4 <- mean(new_df$b_value_pas_4)
c_mean_pas_4 <- mean(new_df$c_value_pas_4)
d_mean_pas_4 <- mean(new_df$d_value_pas_4)
# Calculating mean y_hats pr. pas score
new_df <- new_df %>% 
  group_by(x) %>% 
  mutate(y_hat_1_mean_grand = mean(y_hat_1),
       y_hat_2_mean_grand = mean(y_hat_2),
       y_hat_3_mean_grand = mean(y_hat_3),
       y_hat_4_mean_grand = mean(y_hat_4)) %>% 
  ungroup()

```

```{r}

# Plotting 
new_df %>% 
  ggplot() + 
  geom_smooth(aes(x = x, y = y_hat_1, color = "pas1"), method = "loess") + 
  geom_smooth(aes(x = x, y = y_hat_2, color = "pas2"), method = "loess") + 
  geom_smooth(aes(x = x, y = y_hat_3, color = "pas3"), method = "loess") + 
  geom_smooth(aes(x = x, y = y_hat_4, color = "pas4"), method = "loess") +
  geom_point(aes(x = x, y = y_hat_1_mean_grand, color = "pas1"))+
  geom_point(aes(x = x, y = y_hat_2_mean_grand, color = "pas2"))+
  geom_point(aes(x = x, y = y_hat_3_mean_grand, color = "pas3"))+
  geom_point(aes(x = x, y = y_hat_4_mean_grand, color = "pas4"))+
  labs(title = "Estimated fits for accuracy ratings pr. PAS for all subjects",
       x = "Target.Frames",
       y = "Estimated accuracy ratings using sigmoid-function") +
  theme_bw() 

```

### 7.2.i. compare with the figure you made in 5.3.ii and comment on the differences between the fits - mention some advantages and disadvantages of both.

**ALEKS**: Simply fitting the sigmoid with lowest RSS doesn't take into account the goal of statistics as we see it - it tells us nothing about population, explanatory power, significance etc. Also, by taking the mean of the estimated values, you reduce the resolution of the data, which a glm-model would take into account. Perhaps the sigmoid-by-hand is faster/easier in some cases for plotting or otherwise. *a*
